---
title: "ph_exploratory"
author: "Julius Alipala"
date: "March 6, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r imports}
library(vegan)
```

```{r load}
dat = read.csv('../data/ph_fies.csv')
region_codes = read.csv('../data/region_codes.csv')

```


```{r subset}
# total household income, region, bread/cereals
# rice, meat, seafood, fruit, vegetables
features = c(1,2,3,6,7,8,9,10,11,21,34)

# data with only features to inspect
dat_sub = dat[,features]
```

```{r merge}
dat_merged = merge(dat_sub, region_codes, all.x=TRUE)
```


# vector of regions
region_vector = unique(dat_sub[,2], MARGIN=2)

Luzon = region_vector[c(1, 4, 6, 7, 8, 13, 14, 15)]
Visayas = region_vector[c(3, 9, 12)]
Mindanao = region_vector[c(2, 5, 10, 11, 16, 17)]

island = region_vector[c(Luzon, Visayas, Mindanao)]

test = merge(region_vector, region_codes, by.x='x', by.y='Region', all.x=TRUE)



```{r boxplots}
#par(mar=c(8,4,2,2), cex.axis=0.8)

#png(filename="../figs/rice.png")
boxplot(Total.Rice.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="region", ylab="rice expenditure", las=2)
#dev.off()

#png(filename="../figs/meat.png")
boxplot(Meat.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="region", ylab="meat expenditure", las=2)
#dev.off()

boxplot(Total.Fish.and..marine.products.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="region", ylab="seafood expenditure", las=2)

boxplot(Fruit.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="region", ylab="fruit expenditure", las=2)

boxplot(Vegetables.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="region", ylab="vegetables expenditure", las=2)

#png(filename="../figs/bread.png")
boxplot(Bread.and.Cereals.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="region", ylab="bread/cereals expenditure", las=2)
#dev.off()
```

```{r scatterplot}
plot(Meat.Expenditure ~ Total.Household.Income, data=dat_merged)

points(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, subset= Island == "Luzon",
       col='red')
points(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, subset= Island == "Visayas",
       col='blue')
points(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, subset= Island == "Mindanao",
       col='green')

```



Rice expenditure is quite steady throughout the regions. There is more variation in meat expenditure than I expected. I expected more variation in seafood.


{r mean}
# do not include region in mean calculation
avg_spending = aggregate(dat_merged[,-c(1,10)], list(dat_merged$Region), mean)
dat_expenditures = dat_merged[,-c(1,10)]
mean_expenditures = colMeans(dat_expenditures)
print(avg_spending)
print(mean_expenditures)



On average, citizens spend almost the same on meat and seafood. I'm surprised to see that the average spending on bread/cereals is higher than meat, rice, and seafood.

```{r scatter}
#plot(Meat.Expenditure ~ Total.Household.Income, data= dat_sub, xlab="Total Household Income", ylab="Meat Expenditure")
#pairs(dat_sub)
```

```{r pca}
ph_pca = rda(dat_merged[2:11], scale=TRUE)
ph_pca
```

```{r plot_pca}
plot(ph_pca)
biplot(ph_pca)
#cleanplot.pca(ph_pca)
```

```{r main}
meat_main = lm(Meat.Expenditure ~
               Total.Household.Income + Total.Food.Expenditure + Education.Expenditure + Total.Number.of.Family.members, data=dat_merged)

summary(meat_main)

```

```{r}
income_mod = lm(Meat.Expenditure ~
               Total.Household.Income, data=dat_merged)

summary(income_mod)

island_mod = lm(Meat.Expenditure ~
               Island, data=dat_merged)

summary(island_mod)


```


```{r int}
meat_int = lm(Meat.Expenditure ~
               Total.Household.Income * Island, data=dat_merged)

summary(meat_int)

```

```{r aic}
AIC(meat_int)
AIC(income_mod)
AIC(island_mod)

```

```{r}
par(mfrow = c(2,2))
plot(meat_int)
```


```{r scatter}
# plot main
plot(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, xlab="Total Household Income", ylab="Meat Expenditure")

points(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, subset= Island == "Luzon",
       col='red')
points(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, subset= Island == "Visayas",
       col='blue')
points(Meat.Expenditure ~ Total.Household.Income, data=dat_merged, subset= Island == "Mindanao",
       col='green')

```
