---
title: "ph_exploratory"
author: "Julius Alipala"
date: "March 6, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r imports}
library(vegan)
library(scales)
```

```{r load}
# data from Family Income and Expenditure survey
dat = read.csv('../data/ph_fies.csv')

# data for classifying regions into 3 islands (Luzon, Mindanao, Visayas)
region_codes = read.csv('../data/region_codes.csv')
```


```{r subset}
# features to inspect
features = c("Total.Household.Income", "Region", 
             "Education.Expenditure", "Total.Number.of.Family.members", 
             "Total.Food.Expenditure", "Bread.and.Cereals.Expenditure",
             "Total.Rice.Expenditure", "Meat.Expenditure",
             "Total.Fish.and..marine.products.Expenditure", "Fruit.Expenditure",
             "Vegetables.Expenditure", "Members.with.age.less.than.5.year.old",
             "Members.with.age.5...17.years.old")

# data with only features to inspect
dat_sub = dat[,features]
```

```{r merge}
# add island groups to samples for coarser level of granularity
dat_merged = merge(dat_sub, region_codes, all.x=TRUE)
```

Initial assumptions:
1) Region will not have a significant effect on rice expenditure.
2) Income will not have a significant effect on rice expenditure.
3) People living in Visayas will spend more on seafood.
4) People will spend more on seafood than on meat.
5) Higher income households will spend more on meat.
6) Mindanao will spend less on meat.

Examining income and food expenditures for each island (Luzon, Visayas, Mindanao).
```{r islands}
par(mfrow=c(2,4), mgp = c(3,1,0))

boxplot(Total.Household.Income ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="household income (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Total.Food.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="food expenditure (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Total.Rice.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="rice expenditure (pesos)", 
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Meat.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="meat expenditure (pesos)", 
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Total.Fish.and..marine.products.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="seafood expenditure (pesos)", 
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Fruit.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="fruit expenditure (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Vegetables.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="vegetables expenditure (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)

boxplot(Bread.and.Cereals.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="bread/cereals expenditure (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)
```
Luzon spends more on meat than the other islands, with Mindanao spending the least. Luzon also spends slightly more on fruits and vegetables compared to the other islands. Luzon also has the highest total household income. This seems to support assumptions 1 and 6.


Do people spend more on seafood than on meat?
```{r means}
colMeans(dat_merged[c("Meat.Expenditure", "Total.Fish.and..marine.products.Expenditure")])
```
Assumption 4 does not hold. Meat expenditure is higher on average although it's not by a large amount.


Does Visayas spend more on seafood compared to other islands?
```{r visayas}
aggregate(dat_merged["Total.Fish.and..marine.products.Expenditure"], list(dat_merged$Island), mean)

```
Visayas does spend more on seafood compared to the other islands. This supports assumption 3, although the difference in spending between the islands only differs by a few hundred pesos.


Examining income on food expenditures.
```{r income}
par(mfrow=c(2,4))

plot(Total.Food.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="food expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Total.Food.Expenditure),
      col='red', lwd=2)

plot(Bread.and.Cereals.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="bread/cereals expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Bread.and.Cereals.Expenditure),
      col='red', lwd=2)

plot(Total.Rice.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="rice expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Total.Rice.Expenditure),
      col='red', lwd=2)

plot(Meat.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="meat expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Meat.Expenditure),
      col='red', lwd=2)

plot(Total.Fish.and..marine.products.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="seafood expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income,
             dat_merged$Total.Fish.and..marine.products.Expenditure),
      col='red', lwd=2)

plot(Fruit.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="fruit expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Fruit.Expenditure),
      col='red', lwd=2)

plot(Vegetables.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="vegetables expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Vegetables.Expenditure),
      col='red', lwd=2)

```
The graphs suggest that meat expenditure is affected by total household income and rice expenditure is not highly affected by total household income. Based on this information, assumptions 2 and 5 seem to hold.

Seafood, fruit, and vegetable expenditure also seem to be affected by total household income.
Bread/cereal expenditure is not highly affected


Investigating what factors affect rice expenditures.
Examining other explanatory variables.


Examining effect of number of family members on food expenditure.
```{r family}
par(mfrow=c(2,4))

plot(Total.Household.Income ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="household income", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Total.Household.Income),
      col='red', lwd=2)

plot(Total.Food.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="food expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Total.Food.Expenditure),
      col='red', lwd=2)

plot(Meat.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="meat expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Meat.Expenditure),
      col='red', lwd=2)

plot(Total.Fish.and..marine.products.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="seafood expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Total.Fish.and..marine.products.Expenditure),
      col='red', lwd=2)

plot(Total.Rice.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="rice expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Total.Rice.Expenditure),
      col='red', lwd=2)

plot(Bread.and.Cereals.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="bread and cereal expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Bread.and.Cereals.Expenditure),
      col='red', lwd=2)

plot(Fruit.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="fruit expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Fruit.Expenditure),
      col='red', lwd=2)

plot(Vegetables.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="vegetables expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Vegetables.Expenditure),
      col='red', lwd=2)

```
As the family size increases, spending on food overall also increases. Meat and especially fruit expenditures do not seem to be affected much. An interesting finding is that household income is not affected much by family size. This means that small and large families have about the same income. Fruit may be a lower priority food group, which would explain why there is no noteable increase in fruit expenditure when the family size grows.

Examining effect of education expenditure on food expenditure.
```{r education}
par(mfrow=c(2,4))

plot(Total.Household.Income ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="household income", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Total.Household.Income),
      col='red', lwd=2)

plot(Total.Food.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="food expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Total.Food.Expenditure),
      col='red', lwd=2)

plot(Meat.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="meat expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Meat.Expenditure),
      col='red', lwd=2)

plot(Total.Fish.and..marine.products.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="seafood expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Total.Fish.and..marine.products.Expenditure),
      col='red', lwd=2)

plot(Total.Rice.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="rice expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Total.Rice.Expenditure),
      col='red', lwd=2)

plot(Bread.and.Cereals.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="bread and cereal expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Bread.and.Cereals.Expenditure),
      col='red', lwd=2)

plot(Fruit.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="fruit expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Fruit.Expenditure),
      col='red', lwd=2)

plot(Vegetables.Expenditure ~ Education.Expenditure, data=dat_merged,
        xlab="education", ylab="vegetables expenditure", las=2)
lines(lowess(dat_merged$Education.Expenditure, dat_merged$Vegetables.Expenditure),
      col='red', lwd=2)

```
Spending on education seems to have a substantial affect on household income and food expenditure. Household with higher incomes may be more willing to spend on education. Since households with higher incomes also spend more on food, this may explain why education is also positively correlated with education. A few of the food groups have slight increases as more is spent on education. Rice, bread/cereals, and fruit expenditures seem to be mostly unaffected by spending on education.


Examining if the number of young children affects food expenditures.
```{r children}
par(mfrow=c(2,4))

plot(Total.Household.Income ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="household income", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Total.Household.Income),
      col='red', lwd=2)

plot(Total.Food.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="food expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Total.Food.Expenditure),
      col='red', lwd=2)

plot(Meat.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="meat expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Meat.Expenditure),
      col='red', lwd=2)

plot(Total.Fish.and..marine.products.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="seafood expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Total.Fish.and..marine.products.Expenditure),
      col='red', lwd=2)

plot(Total.Rice.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="rice expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Total.Rice.Expenditure),
      col='red', lwd=2)

plot(Bread.and.Cereals.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="bread and cereal expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Bread.and.Cereals.Expenditure),
      col='red', lwd=2)

plot(Fruit.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="fruit expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Fruit.Expenditure),
      col='red', lwd=2)

plot(Vegetables.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="children less than 5", ylab="vegetables expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Vegetables.Expenditure),
      col='red', lwd=2)

```
Not sure if this is correct. The max number for "Members.with.age.less.than.5.year.old" is 5. While it may not be common, I feel like there would be instances where there would be more than 5 children younger than 5 in a household.

This feature does not seem to have much affect on income or food expenditures.

```{r children_education}
par(mfrow=c(1,2))

plot(Education.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged,
        xlab="age < 5", ylab="education", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Education.Expenditure ),
      col='red', lwd=2)

plot(Education.Expenditure ~ Members.with.age.5...17.years.old, data=dat_merged,
        xlab="5 < age < 17", ylab="education", las=2)
lines(lowess(dat_merged$Members.with.age.5...17.years.old, dat_merged$Education.Expenditure),
      col='red', lwd=2)

```
I was expecting the number of members under the age of 17 to have more affect on the amount spent on education, but this does not seem to be the case.


```{r models}
meat_mod = lm(Meat.Expenditure ~
               log10(Total.Household.Income) + Island, data=dat_merged)
summary(meat_mod)

rice_mod = lm(Total.Rice.Expenditure ~
                Total.Number.of.Family.members, data=dat_merged)
summary(rice_mod)

bread_cereals_mod = lm(Bread.and.Cereals.Expenditure ~
                Total.Number.of.Family.members, data=dat_merged)
summary(bread_cereals_mod)


#termplot(meat_mod, partial.resid = T, se=T)
#termplot(rice_mod, partial.resid = T, se=T)
#termplot(bread_cereals_mod, partial.resid = T, se=T)
```
