---
title: "ph_exploratory"
author: "Julius Alipala"
date: "March 6, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load}
dat = read.csv('../data/ph_fies.csv')
regions = read.csv('../data/regions.csv')

```


```{r subset}
# total household income, region, bread/cereals
# rice, meat, seafood, fruit, vegetables
features = c(1,2,3,6,7,8,9,10,11)

# data with only features to inspect
dat_sub = dat[,features]
```

```{r merge}
dat_merged = merge(dat_sub, regions, all.x=TRUE)
```

```{r}
# vector of regions
region_vector = unique(dat_sub[,2], MARGIN=2)

Luzon = region_vector[c(1, 4, 6, 7, 8, 13, 14, 15)]
Visayas = region_vector[c(3, 9, 12)]
Mindanao = region_vector[c(2, 5, 10, 11, 16, 17)]

island = region_vector[c(Luzon, Visayas, Mindanao)]

test = merge(region_vector, regions, by.x='x', by.y='Region', all.x=TRUE)
```


```{r exploratory}
par(mar=c(8,4,2,2), cex.axis=0.8)

#png(filename="../figs/rice.png")
boxplot(Total.Rice.Expenditure ~ Regione data=dat_sub,
        outline=FALSE, xlab="region", ylab="rice expenditure", las=2)
#dev.off()
```

png(filename="../figs/meat.png")
boxplot(Meat.Expenditure ~ Region, data=dat_sub,
        outline=FALSE, xlab="region", ylab="meat expenditure", las=2)
dev.off()

boxplot(Total.Fish.and..marine.products.Expenditure ~ Region, data=dat_sub,
        outline=FALSE, xlab="region", ylab="seafood expenditure", las=2)

boxplot(Fruit.Expenditure ~ Region, data=dat_sub,
        outline=FALSE, xlab="region", ylab="fruit expenditure", las=2)

boxplot(Vegetables.Expenditure ~ Region, data=dat_sub,
        outline=FALSE, xlab="region", ylab="vegetables expenditure", las=2)

png(filename="../figs/bread.png")
boxplot(Bread.and.Cereals.Expenditure ~ Region, data=dat_sub,
        outline=FALSE, xlab="region", ylab="bread/cereals expenditure", las=2)
dev.off()



Rice expenditure is quite study throughout the regions. There is more variation in meat expenditure than I expected. I expected more variation in seafood.


```{r mean}
# do not include region in mean calculation
avg_spending = aggregate(dat_sub[,-c(2)], list(dat_sub$Region), mean)
dat_expenditures = dat_sub[,-c(2)]
mean_expenditures = colMeans(dat_expenditures)
print(mean_expenditures)
```


On average, citizens spend almost the same on meat and seafood. I'm surprised to see that the average spending on bread/cereals is higher than meat, rice, and seafood.

```{r scatter}
#plot(Meat.Expenditure ~ Total.Household.Income, data= dat_sub, xlab="Total Household Income", ylab="Meat Expenditure")
pairs(dat_sub)
```
