---
title: "ph_models"
author: "Julius Alipala"
date: "April 18, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r imports}
library(vegan)
library(scales)
```
```{r load}
# data from Family Income and Expenditure survey
dat = read.csv('../data/ph_fies.csv')

# data for classifying regions into 3 islands (Luzon, Mindanao, Visayas)
region_codes = read.csv('../data/region_codes.csv')
```


```{r subset}
# features to inspect
features = c("Total.Household.Income", "Region", 
             "Education.Expenditure", "Total.Number.of.Family.members", 
             "Total.Food.Expenditure", "Bread.and.Cereals.Expenditure",
             "Total.Rice.Expenditure", "Meat.Expenditure",
             "Total.Fish.and..marine.products.Expenditure", "Fruit.Expenditure",
             "Vegetables.Expenditure", "Members.with.age.less.than.5.year.old",
             "Members.with.age.5...17.years.old")

# data with only features to inspect
dat_sub = dat[,features]
```
```{r merge}
# add island groups to samples for coarser level of granularity
dat_merged = merge(dat_sub, region_codes, all.x=TRUE)
```

```{r meat plots}
png(filename='../figs/meat_island.png')
boxplot(Meat.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="meat expenditure (pesos)", 
        names=c("Luz", "Min", "Vis"), las=1)
dev.off()

png(filename='../figs/meat_income.png')
plot(Meat.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="meat expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Meat.Expenditure),
      col='red', lwd=2)
dev.off()

```

```{r meat models}
# income and island seem to have the most affect on meat expenditures
meat_int = lm(Meat.Expenditure ~
              log10(Total.Household.Income) * Island, data=dat_merged)

```

```{r meat diagnostics}
summary(meat_int)

par(mfrow=c(2,2))
termplot(meat_int, partial.resid = T, se=T)
```

```{r rice plots}

png(filename='../figs/rice_fsize.png')
plot(Total.Rice.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="rice expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Total.Rice.Expenditure),
      col='red', lwd=2)
dev.off()

png(filename='../figs/rice_children.png')
boxplot(Total.Rice.Expenditure ~ Members.with.age.less.than.5.year.old, data=dat_merged, 
        outline=FALSE, xlab="children less than 5", ylab="rice expenditure", las=2)
lines(lowess(dat_merged$Members.with.age.less.than.5.year.old, dat_merged$Total.Rice.Expenditure),
      col='red', lwd=2)

dev.off()

```

```{r rice models}
# family size and number of young children have most affect on rice expenditure
rice_int = lm(Total.Rice.Expenditure ~ 
               Total.Number.of.Family.members * Members.with.age.less.than.5.year.old,  
              data=dat_merged)

```
```{r rice diagnostics}
summary(rice_int)
par(mfrow=c(2,2))
termplot(rice_int, partial.resid = T, se=T)
```
Number of young children has less affect on rice expenditure than I thought. Interaction affect model seems to be better.


```{r bread plots}

png(filename='../figs/bread_income.png')
plot(Bread.and.Cereals.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="bread/cereals expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Bread.and.Cereals.Expenditure),
      col='red', lwd=2)
dev.off()

png(filename='../figs/bread_fsize.png')
plot(Bread.and.Cereals.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="bread and cereal expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Bread.and.Cereals.Expenditure),
      col='red', lwd=2)
dev.off()


```


```{r bread models}
# income, family size, and number of young children have most affect on bread/cereal expenditure
bread_int = lm(Bread.and.Cereals.Expenditure ~ 
               Total.Number.of.Family.members * log10(Total.Household.Income),  
              data=dat_merged)

```
```{r bread diagnostics}
summary(bread_int)

par(mfrow=c(2,2))
termplot(bread_int, partial.resid = T, se=T)
```
Similar to the examination of rice expenditures, number of young children doesn't seem to be a big factor in bread/cereal expenditure. Income has a significant affect on bread/cereal expenditure. Interaction affects model is slightly better. 


```{r seafood plots}

png(filename='../figs/seafood_income.png')
plot(Total.Fish.and..marine.products.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="seafood expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income,
             dat_merged$Total.Fish.and..marine.products.Expenditure),
      col='red', lwd=2)
dev.off()

png(filename='../figs/seafood_fsize.png')
plot(Total.Fish.and..marine.products.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="seafood expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Total.Fish.and..marine.products.Expenditure),
      col='red', lwd=2)
dev.off()

png(filename='../figs/seafood_island.png')
boxplot(Total.Fish.and..marine.products.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="seafood expenditure (pesos)", 
        names=c("Luz", "Min", "Vis"), las=1)
dev.off()



```


```{r seafood models}
# income, family size, and island seem to have the most affect on seafood expenditures
seafood_int = lm(Total.Fish.and..marine.products.Expenditure ~
              log10(Total.Household.Income) * Island * Total.Number.of.Family.members,
              data=dat_merged)


```

```{r seafood diagnostics}
summary(seafood_int)

par(mfrow=c(2,2))
termplot(seafood_int, partial.resid = T, se=T)
```
Interaction affect is the best model. The termplot for number of family members in interaction affect model is interesting.

```{r fruit plots}

png(filename='../figs/fruit_island.png')
boxplot(Fruit.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="fruit expenditure (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)
dev.off()

png(filename='../figs/fruit_income.png')
plot(Fruit.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="fruit expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Fruit.Expenditure),
      col='red', lwd=2)
dev.off()


```



```{r fruit models}
# island and income seems to have the most affect on fruit expenditure
fruit_int = lm(Fruit.Expenditure ~ 
               log10(Total.Household.Income) * Island,  
              data=dat_merged)

```
```{r fruit diagnostics}
summary(fruit_int)

par(mfrow=c(2,2))
termplot(fruit_int, partial.resid = T, se=T)
```

Interaction effect model is the best. Island doesn't seem to have much affect. Household income is a better indicator of fruit expenditures.


```{r vegetable plots}

png(filename='../figs/veg_income.png')
plot(Vegetables.Expenditure ~ Total.Household.Income, data=dat_merged,
     xlab="household income (pesos)", ylab="vegetables expenditure (pesos)", las=1)
lines(lowess(dat_merged$Total.Household.Income, dat_merged$Vegetables.Expenditure),
      col='red', lwd=2)
dev.off()

png(filename='../figs/veg_island.png')
boxplot(Vegetables.Expenditure ~ Island, data=dat_merged,
        outline=FALSE, xlab="island", ylab="vegetables expenditure (pesos)",
        names=c("Luz", "Min", "Vis"), las=1)
dev.off()

png(filename='../figs/veg_fsize.png')
plot(Vegetables.Expenditure ~ Total.Number.of.Family.members, data=dat_merged,
        xlab="family size", ylab="vegetables expenditure", las=2)
lines(lowess(dat_merged$Total.Number.of.Family.members, dat_merged$Vegetables.Expenditure),
      col='red', lwd=2)
dev.off()

```


```{r vegetable models}
# island, family size and income seems to have the most affect on vegetable expenditure
vegetable_int = lm(Vegetables.Expenditure ~ 
               log10(Total.Household.Income) * Island * Total.Number.of.Family.members,  
              data=dat_merged)

```
```{r vegetable diagnostics}
summary(vegetable_int)

par(mfrow=c(2,2))
termplot(vegetable_int, partial.resid = T, se=T)
```
Interaction affect is the better model. Income has the biggest affect. Family size has a much smaller affect and island has a very small affect.



